---
title: "03-Data-Visualisation"
output:
  html_notebook:
    toc: yes
---
```{r label = "global-options", echo=FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA)
```

# 3.1 Prerequisites

```{r label="Prerequisites"}
library(tidyverse)
# if you need updates, run 
# tidyverse_update()
# make the folder, where this file originate to the working directory
setwd("~/Documents/_PB-Data/Programming/R/Learning-R/Coursera/R4ds-book")
# this does not work, maybe the working directory is determined by the project?
# load all tidyverse packages

```
# 3.2 First steps
## 3.2.1 The `mpg` data frame
```{r}
mpg
class(mpg)
head(mpg)
tail(mpg)
str(mpg)
```
## 3.2.2 Creating a ggplot
```{r label="creating-a-ggplot"}
ggplot(data = mpg) + 
        geom_point(mapping = aes(x = displ, y = hwy))
```
You can make a template, where you just have to substitute the expressions with brackets with your data.
## 3.2.3 A graphic template
```{r label="a-graphic-template"}
# ggplot(data = <DATA>) + 
#         <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
## 3.2.4 Exercises

(1) Run `ggplot(data = mpg)` -- what do you see?
```{r}
ggplot(data = mpg) # an empty graphic frame
```


(2) How many rows are in mtcars? How many columns? `r nrow(mpg)` rows and `r ncol(mpg)` columns.

(3) What does the drv variable describe? Read the help for ?mpg to find out.
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

(4) Make a scatterplot of hwy vs cyl.
```{r label="scatterplot-of-hwy-vs-cyl"}
ggplot(data = mpg) + 
         geom_point(mapping = aes(x = hwy, y = cyl))
```


(5) What happens if you make a scatterplot of class vs drv. Why is the plot not useful?
```{r}
ggplot(data = mpg) + 
          geom_point(mapping = aes(x = class, y = drv))
```

