---
title: "Test with Umlaute"
output: 
    github_document:
            toc: yes
            toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.


## Including Code

You can include R code in the document as follows:



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
x <- "Datens채tze"
cat(stringi::stri_escape_unicode(x))
```

Es gibt kein Problem innerhalb des Textes mit Datens채tze. Wie ist es mit Links von [Datens채tze](https://en.wikipedia.org/wiki/ISO_639-1) zu einer Website?

## Datens채tzen

Here comes new text.
